### 2013-09-30 Publications/2013-ISBI14/01-abstract.tex
Kieran's comments and some other improvements

\begin{abstract}
Connectivity analysis on diffusion MRI data of the whole-brain
suffers from distortions originated by the standard 
echo-planar imaging sequences used in acquisition. %
%These artifacts are present at interfaces between tissues,
%due to the inhomogeneity of the field derived from the 
% discontinuity of magnetic susceptibility.
These images show characteristic geometrical deformations
and signal destruction that suppose a significant pitfall
limiting the success of tractography algorithms.

>>>>

\begin{abstract}
Connectivity analysis on diffusion MRI data of the whole-brain
suffers from distortions originated by the standard 
echo-planar imaging sequences used in acquisition. %
%These artifacts are present at interfaces between tissues,
%due to the inhomogeneity of the field derived from the 
% discontinuity of magnetic susceptibility.
These images show characteristic geometrical deformations
and signal destruction that are a significant pitfall
limiting the success of tractography algorithms.

### 2013-09-30 Publications/2013-ISBI14/02-introduction.tex
Kieran's comments and some other improvements

\section{INTRODUCTION}

In-vivo whole-brain connectivity analysis has been a
research topic of high interest for the last
5 years. 
\Gls*{dmri} can be used to probe the
orientation of fiber bundles within the brain,
generally applying \gls*{epi} sequences.
After a signal reconstruction step, 
tractography algorithms draw a map of the sampled 
structures.
These maps can represent the actual trajectories
of fiber bundles (deterministic tractography) or
pixel-wise probability of connection to a certain origin
(probabilistic tractography). Finally, the
information about these connections is collected
into a network matrix that can be subjected to
the so-called ``connectome analysis''.

Among all the difficulties that such a complex workflow
raises \cite{jones_twenty-five_2010}, we will
address here the susceptibility-derived artifacts,
for which \Gls*{epi} schemes are specially sensitive.
As susceptibility changes at tissue interfaces,
so does the magnetic field. This inhomogeneities 
of the field translate in a highly distorted imaged 
anatomy and a significant signal  destruction of 
certain regions of the brain 
(e.g. the orbitofrontal lobe, for the proximity of the
air surrounding sinuses). This artifact has been
profoundly described, generally within the context of
functional MRI because it is usually acquired with 
\gls*{epi} as \gls*{dmri}.

\Gls*{fmb} methods supposed the first approach to retrospective
correction \cite{jezzard_correction_1995}. They rely on one
extra acquisition (\emph{field mapping}), that probes the 
inhomogeneity of the field.
A second theory-based breed of methodologies acquire a 
map of the \acrlong*{psf} of the \gls*{epi} readouts to correct 
for the artifact \cite{robson_measurement_1997}. 
Next generation of methodologies \cite{cordes_geometric_2000,
chiou_simple_2000}, make use of the acquisition of 
extra \gls*{epi} volumes 
with specific differences that enable correction, for instance
the opposed direction of the phase encoding gradients. We will
refer to these methods as \gls*{reb}.
Finally, the last family of methodologies acquire an 
extra T2-weighted image, and use its anatomical correctness 
as reference to find the deformation map through nonlinear 
registration \cite{kybic_unwarping_2000,studholme_accurate_2000}.
\gls*{t2b} methods usually map
the T2 image to the \emph{baseline} volume or ``B0''
of \gls*{dmri}. The choice of T2 is due to the strong
similarity of intensity distribution with respect the B0.
More recent works report extensions or combined
approaches of existing techniques
\cite{andersson_how_2003,zaitsev_point_2004,%
holland_efficient_2010,andersson_comprehensive_2012}.

>>>>

\section{INTRODUCTION}
\label{sec:intro}
In-vivo whole-brain connectivity analysis has been a
research topic of high interest for the last
5 years. 
\Gls*{dmri} can be used to probe the
orientation of fiber bundles within the brain,
and is generally acquired with an \gls*{epi} sequence.
After a signal reconstruction step, 
tractography algorithms draw a map of the sampled 
structures.
These maps can represent the actual trajectories
of fiber bundles (deterministic tractography) or
pixel-wise probability of connection to a certain origin
(probabilistic tractography). Finally, the
information about these connections is collected
into a network matrix that can be subjected to
the so-called ``connectome analysis''.

Among all the difficulties that such a complex workflow
raises \cite{jones_twenty-five_2010}, we will
address here the susceptibility-derived artifacts,
for which \Gls*{epi} schemes are especially sensitive.
Magnetic susceptibility disturbs the magnetic field 
close to tissue interfaces. This inhomogeneity 
of the field translates in a highly distorted imaged 
anatomy and a significant signal destruction of 
certain regions of the brain 
(e.g. the orbitofrontal lobe, for the proximity of the
air surrounding sinuses). This artifact has been
profoundly described, generally within the context of
functional MRI because they are also acquired with 
\gls*{epi}.

\Gls*{fmb} methods supposed the first approach to retrospective
correction \cite{jezzard_correction_1995}. They rely on one
extra acquisition (\emph{field mapping}), that probes the 
inhomogeneity of the B0 field.
A second theory-based breed of methodologies acquire a 
map of the \acrlong*{psf} of the \gls*{epi} readouts to correct 
for the artifact \cite{robson_measurement_1997}. 
Next generation of methodologies \cite{cordes_geometric_2000,
chiou_simple_2000}, make use of the acquisition of 
extra \gls*{epi} volumes 
with specific differences that enable correction, for instance
the opposed direction of the phase encoding gradients. We will
refer to these methods as \gls*{reb}.
Finally, the last family of methodologies acquire an 
extra T2-weighted image, and use its anatomical correctness 
as reference to find the deformation map through nonlinear 
registration \cite{kybic_unwarping_2000,studholme_accurate_2000}.
\gls*{t2b} methods usually map
the T2 image to the \emph{baseline} volume or \textit{b0}
of \gls*{dmri}. The choice of T2 is due to the strong
similarity of intensity distribution with respect the \textit{b0}.
More recent works report extensions or combined
approaches of existing techniques
\cite{andersson_how_2003,zaitsev_point_2004,%
holland_efficient_2010,andersson_comprehensive_2012}.

### 2013-09-30 Publications/2013-ISBI14/03-methods.tex
Kieran's comments and some other improvements

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the High angular resolution diffusion imaging 
reconstruction Challenge held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as a \gls*{dmri}
(1.0mm isotropic, $b$=1200, 32 evenly-distributed 
directions, 1 B0 image) image.
The phantom includes \gls*{wm} fiber bundles, 
\gls*{csf} and \gls*{gm}. Diffusion is modeled by a 
restricted and a hindered compartment, similar to
\cite{assaf_composite_2005}.

\subsection{Theory-based synthetic distortion}

\Gls*{fmb} methodologies use a map
of the field in the scanner. More precisely, the
phase difference between two subsequent samplings
of the fieldmap. With that information, it is possible
to compute the theoretical displacement that each
voxel is to suffer, the so-called \gls*{vsm}. The
most prominent feature of this \gls*{vsm} is that all
the shifts have the same orientation (parallel to the
phase-encoding direction of the \gls*{epi}) and their
magnitude and direction depend on the \gls*{epi} 
gradient increments (or \emph{blips}), and the actual
phase difference at the voxel.

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom.  We defined two regions 
of smooth de-phasing (up to $0.05\pi$ rad. excursions) 
and computed the corresponding \gls*{vsm}, using the 
tools distributed with the \texttt{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46ms for the
fieldmap and \emph{effective dwell time} of 
0.77ms for the \gls*{epi}).

@@

\begin{figure}[thpb]
   \centering
   \includegraphics[width=0.8\columnwidth]{Fig01-Phantom}
   \caption{Original and distorted digital phantom (sagittal
   planes). A) T2-weighted image; B) undistorted B0 volume;
   C and D) distorted B0 volumes with opposed phase encoding 
   directions.}
   \label{fig:label}
\end{figure}

@@

\paragraph*{\Gls*{fmb}} As it was previously described,
this family of correction technologies infer a \gls*{vsm}
from the phase difference map. In this work, we use the
tool \texttt{fugue} \cite{jenkinson_fsl_2012} aforementioned.

\paragraph*{\Gls*{reb}} We use the tool
\texttt{topup} from \cite{jenkinson_fsl_2012} to correct
using the B0 of a reverse-encoding simulation of the \gls*{dti}.
In this second case, the \gls*{vsm} is inferred from the
differences between the two corresponding B0. This technique
also uses least-squares-fitting to interpolate the undistorted
signal, and therefore provides with a robust dropout correction.

\paragraph*{\Gls*{t2b}} For this last methodology
we fine-tuned the \texttt{ANTs} tool \cite{avants_ants:_2013},
in order that registering the B0 to the T2. Additionally,
undistorted images are corrected for dropout using the determinant
of the Jacobian of the deformation field.

@@

\paragraph*{Proposed workflows}
We used \texttt{nipype} \cite{gorgolewski_nipype:_2011}
to conveniently integrate the correction methods in
comprehensive evaluation workflows:

\begin{itemize}
\item \textbf{Phantom deformation}: We release the pipeline
that generates the gold-standard, along with
distorted versions of the \glspl*{tpm} for
evaluation purposes.
\item \textbf{Correction workflow}: A high-level pipeline
integrates the three proposed correction methods,
ensuring that all the requirements for inputs are
met. As outputs this workflow produces the different
corrected versions of the phantom (\gls*{fmb}, \gls*{reb}
and \gls*{t2b}) alongside corrected versions of the
distorted \glspl*{tpm}.
\item \textbf{Reconstruction workflow}: This pipeline uses
the Diffusion Toolkit \cite{wang_diffusion_2007}, 
to produce deterministic tractography
from \gls*{dmri} data at the input. Therefore,
tractographies of original, and recovered data
are produced for comparison.
\item \textbf{Benchmarking workflow}: The final processing
element collects all the outputs from the previous
workflows and extracts the evaluation scores.
\end{itemize}
General-purpose 
workflows have been contributed to the upstream code
of the tool, under the branch \texttt{nipype.workflows.dmri}.
Testing-specific workflows are also publicly released.

\paragraph*{Evaluation scores}
We evaluated three characteristics of the corrected
phantom. First, we assessed the geometrical correctness
and reported overlap indices of three \glspl*{tpm} 
(namely \gls*{csf},\gls*{wm}, and \gls*{gm}).
More precisely, we computed the \gls*{fji} as 
proposed in \cite{crum_generalized_2006} weighted by 
the volumes of labels. 
Second, to evaluate the quality of the actual 
signal \emph{sanity} after recovery (dropout 
correction), we studied the signal similarity
frame by frame computing the $\ell_1$-norm
based correlation index. We report this score
on the B0 and the average of the remaining 32 
\glspl*{dwi}. Third, we report the 
impact on tractography using the \gls*{frr}, that 
informs the number of recovered fibers versus the 
gold-standard number for a specific reconstruction
method.

>>>>

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the High angular resolution diffusion imaging 
reconstruction Challenge held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as a \gls*{dmri}
(1.0mm isotropic, $b$=1200, 32 evenly-distributed 
directions, 1 \textit{b0} image) image.
The phantom includes \gls*{wm} fiber bundles, 
\gls*{csf} and \gls*{gm}. Diffusion is modeled by a 
restricted and a hindered compartment, similar to
\cite{assaf_composite_2005}.

\subsection{Theory-based synthetic distortion}
\label{sec:distortion}
\Gls*{fmb} methodologies use a map
of the field in the scanner. More precisely, the
phase difference between two subsequent samplings
of the fieldmap. With that information, it is possible
to compute the theoretical displacement that each
voxel undergoes, the so-called \gls*{vsm}. The
most prominent feature of this \gls*{vsm} is that all
the shifts have the same orientation (parallel to the
phase-encoding direction of the \gls*{epi}) and their
magnitude and direction depend on the \gls*{epi} 
gradient increments (or \emph{blips}), and the actual
phase difference at the voxel.

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom.  We defined two regions 
of smooth de-phasing (up to $0.05\pi$ rad. excursions) 
and computed the corresponding \gls*{vsm}, using the 
tools distributed with the \emph{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46ms for the
fieldmap and \emph{effective dwell time} of 
0.77ms for the \gls*{epi}).

@@

\begin{figure}[thpb]
   \centering
   \includegraphics[width=0.95\columnwidth]{Fig01-Phantom}
   \caption{Original and distorted digital phantom.
   A) T2-weighted image; B) undistorted \textit{b0} volume;
   C, D) distorted \textit{b0} volumes with opposed phase encoding 
   directions.}
   \label{fig:label}
\end{figure}

@@

Three out of four methods presented in \autoref{sec:intro}
were tested on the evaluation framework.

\paragraph*{\Gls*{fmb}} In this case, the pipeline essentially
the same as the one used for generating the distortion, but
applying the \gls*{vsm} in the opposite direction.

\paragraph*{\Gls*{reb}} We made use of the implementation found
in \emph{FSL} (called \texttt{topup}) to correct using the \textit{b0} of a
reverse-encoding simulation of the \glspl*{dwi}. In this second
case, a \gls*{vsm} is inferred from the differences between the 
two corresponding \textit{b0}.

\paragraph*{\Gls*{t2b}} For the last surveyed method,
we fine-tuned \emph{ANTs} \cite{avants_ants:_2013},
in order to register \textit{b0} to T2. To this end, we used a
multi-resolution scheme with 3 levels of subsampling and smoothing,
mutual information metric, and symmetric diffeomorphic transform 
(\texttt{SyN}). Several configurations of kernel widths for the 
regularization smoothers were tested, and finally selected 
0.5/1.0 voxels (gradient/deformation fields, respectively) 
for its best result. Additionally, undistorted images are 
corrected for dropout using the determinant
of the Jacobian of the deformation field.

@@

The four modules described so far (namely, phantom
deformation, \gls*{fmb} correction, \gls*{reb}
correction and \gls*{t2b} correction) were conveniently
integrated in second order workflows using
\emph{nipype} \cite{gorgolewski_nipype:_2011}.
The choice of this tool grants the reproducibility
of the experiments, among other advantages.

In order to enable the analysis of the influence of
distortions on the final connectivity analyses,
a intermediate step was also implemented to
provide with deterministic tractography,
using the \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}.
The original phantom, one distorted version, and
the corrected instances are fed to this module.
Additionally, the original tissue probability maps
are also distorted and corrected, for computing overlap
indices.
Each tractography was restricted by its corresponding 
\gls*{wm} mask (e.g. the distorted and corrected with
fieldmap for the tractography that followed this method). 
In all cases, we used 10 random seeds at each seeding
voxel and the seeding voxels were taken from the 
ground-truth, mimicking the procedure with real data
(in which this regions are typically projected from the
anatomically correct T1).

Finally, the proposed workflows are connected to performance
assessment modules. We evaluated three characteristics
of the correction methods. 
First, we assessed the geometrical correctness
and reported overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}),
weighting the average by the volumes of labels.
Second, to evaluate the quality of the actual 
signal dropout correction, we studied the 
signal similarity frame by frame computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining 32 
\glspl*{dwi}. Third, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

### 2013-09-30 Publications/2013-ISBI14/04-results.tex
Kieran's comments and some other improvements

\section{RESULTS AND DISCUSSION}

WIP: work in progess..

>>>>

\section{RESULTS AND DISCUSSION}
\label{sec:results}
The results of the proposed experiments are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of results.

\input{041-results-table}

\begin{figure*}[tpb]
   \centering
   \includegraphics[width=0.9\textwidth]{Fig02-Results}
   \caption{The three major stages of the evaluation framework,
   along with the associated visual results. First row represents
   a coronal section of the \textit{b0} volume. In second row, the outcome
   of tractography, showing only tracks that connect two different
   network nodes. Third row shows the associated connectivity matrix. }
   \label{fig:results}
\end{figure*}

\subsection{Geometrical correctness}

The three methods under study achieved good results in
terms of geometrical accuracy, as it is shown in the first
row of \autoref{fig:results}. The best numerical scores
were obtained with the \gls*{reb} method (\autoref{table:results}),
that achieved a weighted average overlap of $96.64\%$. Second
best were obtained with \gls*{fmb}. Geometrical correctness
is fundamental in connectivity analysis to spatially locate the
\glspl*{roi} that will define the nodes of the final connectivity
matrix. The clear difference of accuracy between \gls*{reb} and \gls*{fmb}
with respect to \gls*{t2b} infers the latter may not be
an appropriate method for susceptibility correction.

\subsection{Signal loss recovery}

A second workflow evaluated the different correction 
methodologies to show the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed significantly
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining 31 \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirm the 
results presented in the Table (\autoref{fig:results}, first
row). Finally, \gls*{t2b} confirmed that its correction
can not be compared, even when using the intensity compensation
based on the determinant of the deformation field.

\subsection{Impact on tractography and connectivity}

The gold-standard contained a total of 1800 valid tracks
connecting regions, with an average length of $40.88\pm8.06$mm.
The corresponding connectivity matrix presented 25 connections 
between the total of 27 different seeding regions.

The distortion of the phantom caused 2 false positives (inexistent 
connections created by distortion) and 2 false negatives 
(connections vanished). The correction methods had an adverse impact
on connectivity. Only \gls*{reb} could recover one lost connection
but increasing the false connections count to 3.
\gls*{fmb} ranked second, with 5 false and 2 lost connections.
Finally, \gls*{t2b} presented 9 false and 3 lost connections.

For the shake of completeness, we also document the outcome
of tractography after each correction method. After distortion,
we obtained 1750 tracks and a characteristic length of 
$40.64\pm8.12$mm.
\gls*{fmb} produced 1900 tracks/$40.53\pm8.07$mm., \gls*{reb}
obtained 2200 tracks/$41.15\pm7.60$mm., and 2500 
tracks/$40.60\pm7.81$mm. for \gls*{t2b}. Although track counting has
been used as a measurement of success and as a weight of connectivity 
strength, the number of tracks produced by tractography strongly rely
on the seeding performed and the smoothing caused by interpolations 
within correction methodologies.

\subsection{Discussion}

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate methodology for correction of susceptibility-induced
distortions in \gls*{dwi}. Moreover, \gls*{reb} also achieved
the best results in terms of connectivity, being able to recover
one of two lost connections after distortion. \Gls*{fmb} achieved 
very positive scores. Finally, the \gls*{t2b} method did not achieve
the necessary high-standards to recommend its use. We understand
that specific methods with anisotropic regularization that completely
restrict deformations to the phase-encoding axis would perform
significantly better as compares to the standard methods in this study.

One relevant limitation of this work is that the \gls*{fmb} is
used both in distortion creation and correction, what could
bias results. Nonetheless, \gls*{reb} performed better that \gls*{fmb}.
A second limitation is found on the impossibility of integrating the
point spread function-based methods for two reasons.
First, the unmet need of a point spread function map consistent
with the synthetic phase-difference map. Second, to our knowledge,
there are no implementations of the method readily available.
In practice, this second limitation is narrowed down by the
fact that typical acquisition protocols usually include one or more
datasets among field mapping, reversed-encoding \textit{b0} or T2,
but they do not include the point-spread function mapping.
